Lost in the woods: shifting habitats can lead phylogeography astray

### 2025-03-24 manuscript/ResponseToReviews.tex

Phylogeography aims to embed phylogenetic trees in geographic space to infer past locations and movements. This manuscript critically examines the assumptions underlying the most influential phylogeographic inference methods and assesses their robustness. Neher demonstrates that a widely used summary statistic, “lineage speed,” derived from diffusion-based models, is fundamentally flawed as it depends on sample size. More critically, under realistic demographic scenarios—such as expanding populations or those not strictly governed by local density regulation—lineages exhibit strong backward-in-time biases, despite unbiased forward-time movements. This discrepancy can result in biased and overconfident phylogeographic inferences, which often assume unbiased lineage movement backward in time. Based on these findings, Neher concludes that the validity of phylogeographic inference methods is difficult to assess and should not be trusted without additional information, such as detailed time-series data.

This paper offers an important and long-overdue critique of phylogeographic methods, which have gained significant popularity due to their visual appeal—a popularity that, in light of this critique, appears questionable. The manuscript is exceptionally well-written and pedagogical, making the limitations of diffusion-based methods strikingly clear. The discussion is both illuminating and practical, with helpful examples.

\response{I thank the reviewer for this positive assessment.}

1.  ``This model assumes that the dispersal of one lineage is independent of other lineages, that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time. Furthermore, it assumes that the tree topology and branching rates are independent of spatial location.'' I agree with these points, but it would be helpful to explain why these assumptions are implied by the seemingly innocuous Eq. 1. Users of diffusion-based phylogeographic inference methods may not be aware that these assumptions are inherent.

\response{Thank you for this suggestion. I have expanded this section.}

2. The section on density regulation seems to assume global or habitat-wide density regulation. However, it is plausible that many species are density-regulated over a characteristic length scale, such as their dispersal distance during their lifetime (or from birth to reproductive age). In that case, wouldn’t the parameter  L  (habitat size) be replaced with the scale over which density is regulated? If so, this should be mentioned. A static, homogeneous, and locally density-regulated population should then approximately follow Eq. 1.

\response{TODO}

3. ``which cab lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low'': Replace ``cab'' with ``can''. Additionally, in this context, beyond Wilkins and Wakeley, it may be worthwhile to cite the Appendix B of Hallatschek, O., & Nelson, D. R. (2008). Gene surfing in expanding populations. Theoretical population biology, 73(1), 158-170.

\response{Done.}

4. It might be helpful to provide guidance on whether users can make quantitative estimates or perform tests to assess the reliability of their phylogeographic inferences. For example, how dense does a time series need to be to sufficiently constrain the spatiotemporal embedding? Alternatively, this manuscript could call for studies that address this issue. Following this work, the onus is on the phylogeographic community to develop best practices for these methods.

\response{TODO}

\end{document}
